Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2017-05-03 manuscript.tex
latest

\begin{figure}
\centering
\begin{tabular}{l*{4}{c}}
  & Ideal & Disturbance* & Stress (G) & Stress (HD) \\ \hline
  Constraints & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
  Other parameters & $b/d\gg 1$ & $(1-d)b\approx d$ & $b\approx d$ & $b/d\gg 1$ \\
  Propagule density $L$  & $\gg 1$ & $\ll 1$ & $\ll 1$ & $\gg 1$
\end{tabular}
\caption{\label{fig:table} The mathematical realization of Grime's three environmental extremes in our model, as well as the high-density (HD) variant of the stressful environment. *Mortality affects both adults and juveniles in the disturbed environment, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$.}
\end{figure}

The expected direction of evolution is determined by three factors. First, there may be pleiotropy between $b$, $c$, and $d$, which could be due to trade-offs between them. Second, adaptive mutations in one trait may occur more frequently than in others, which can change the direction of evolution \citep{yampolsky_2001}. Quantifying how traits are affected by mutations is beyond the scope of this article, and so we will not evaluate either of these effects: mutations affect $b$, $c$ and $d$ independently, and they all have the same distribution of beneficial fitness effects.

The third effect is that selection can favor some trait changes over others. The probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}. Very few neutral mutations fix (probability of order $1/N$). Thus $b$, $c$ and $d$ will tend to change in the direction of greatest fitness increase.

Under Eq. \eqref{eq:master}, proportional $b$- and $d$-mutations of the form $b_j=(1+\epsilon) b_i$ and $d_j=(1-\epsilon) d_i$ have identical fitness effects (here $j$ is the mutant type; Appendix C). When propagule density is high  $L\gg 1$, the same is true for $c$-mutations of the form $c_j=(1+\epsilon) c_i$. When propagule density is low $L\ll 1$, $c$ does not contribute to fitness and $c$-mutations are neutral. Thus, in ideal and stressed environments, selection under Eq. \eqref{eq:master} does not intrinsically favor one trait improvement over another, apart from the irrelevance of $c$ at low density.

The situation in disturbed environments is more complicated. The bulk of the high mortality $d\approx 1$ is environmental and unavoidable. Thus, mutational changes in $d$ are intrinsically less effective than mutation change in $b$. However, this effect is easily swamped by the fact that disturbance mortality affects both juvenile recruitment and adult mortality.

>>>>

\begin{figure}
\centering
\begin{tabular}{l*{4}{c}}
  & Ideal & Disturbance* & Stress (G) & Stress (HD) \\ \hline
  Constraints & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
  Other parameters & $b/d\gg 1$ & $(1-d)b\approx d$ & $b\approx d$ & $b/d\gg 1$ \\
  Propagule density $L$  & $\gg 1$ & $\ll 1$ & $\ll 1$ & $\gg 1$
\end{tabular}
\caption{\label{fig:table} The mathematical representation of Grime's three environmental extremes in our model, as well as the high-density (HD) variant of the stressful environment. *Mortality affects both adults and juveniles in the disturbed environment, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$.}
\end{figure}

The expected direction of evolution is determined by three factors. First, there may be pleiotropy between $b$, $c$, and $d$, which could be due to trade-offs between them. Second, adaptive mutations in one trait may occur more frequently than in others \citep{yampolsky_2001}. We do no doubt that these factors operate in nature and contribute towards ``primary strategy'' trait patterns. However, modeling how mutations affect traits subject to trade-offs is beyond the scope of this article, and so we will not evaluate these first two factors: we assume that mutations affect $b$, $c$ and $d$ independently, and that $b$, $c$ and $d$ have the same distribution of beneficial fitness effects.

The third factor is that some traits may have a greater impact on fitness: $b$, $c$ and $d$ will tend to evolve in the direction of greatest fitness increase because the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}. Very few neutral mutations fix (probability of order $1/N$).

Under Eq. \eqref{eq:master}, mutations which change $b$ or $d$ by the same  proportion (i.e. $b_j=(1+\epsilon) b_i$ or $d_j=(1-\epsilon) d_i$ where $j$ is the mutant type) have identical fitness effects (Appendix C). When propagule density is high  $L\gg 1$, the same is true for proportional $c$-mutations $c_j=(1+\epsilon) c_i$. When propagule density is low $L\ll 1$, $c$ does not contribute to fitness and $c$-mutations are neutral. Thus, selection under Eq. \eqref{eq:master}, which applies in ideal and stressed environments, does not intrinsically favor one trait improvement over another, apart from the irrelevance of $c$ at low density.

The situation in disturbed environments, where mortality also affects juveniles, is more complicated. The bulk of the high mortality $d\approx 1$ is environmental and unavoidable. Thus, proportional changes in the remaining genetic part of mortality can have a tiny impact on $d$, suggesting the primacy of $b$ under disturbance. However, this effect is easily swamped by the fact that disturbance mortality affects both juvenile recruitment and adult mortality.
